Generalized Plackett-Luce Likelihoods

### 2024-05-08 inst/paper.tex
Add JSS's paper.tex as per issue #153

@@

%% an abstract and keywords
\Abstract{

  The \code{hyper2} package provides functionality to work with
  extensions of the Bradley-Terry probability model such as
  Plackett-Luce likelihood including team strengths and reified
  entities (monsters).  The package allows one to use relatively
  natural R idiom to manipulate such likelihood functions.  Here, I
  present a generalization of \code{hyper2} in which multiple entities
  are constrained to have identical Bradley-Terry strengths.  A new
  \proglang{S3} class \code{hyper3}, along with associated methods, is
  motivated and introduced.  Three datasets are analysed, each
  analysis furnishing new insight, and each highlighting different
  capabilities of the package.  } \Keywords{Plackett-Luce,
  Bradley-Terry, Mann-Whitney} \Plainkeywords{Plackett-Luce,
  Bradley-Terry, Mann-Whitney}

@@

The \code{hyper2} package~\citep{hankin2009,hankin2017} furnishes
computational support for generalized
Plackett-Luce~\citep{plackett1975} likelihood functions.  The
preferred interpretation is a race (as in track and field athletics):
Given six competitors $1-6$, we ascribe to them nonnegative strengths
$p_1,\ldots, p_6$; the probability that $i$ beats $j$ is
$p_i/(p_i+p_j)$.  It is conventional to normalise so that the total
strength is unity, and to identify a competitor with his strength.
Given an order statistic, say $p_1\succ p_2\succ p_3\succ p_4$, the
Plackett-Luce likelihood function would be

\begin{equation}\label{PL_like}
  \frac{p_1}{p_1+p_2+p_3+p_4}\cdot
  \frac{p_2}{    p_2+p_3+p_4}\cdot
  \frac{p_3}{        p_3+p_4}\cdot
  \frac{p_4}{            p_4}.
\end{equation}

\citet{mollica2014} call this a forward ranking process on the grounds
that the best (preferred; fastest; chosen) entities are identified in
the same sequence as their rank.

Computational support for Bradley-Terry likelihood functions is
available in a range of languages.  \cite{hunter2004}, for example,
presents results in \proglang{MATLAB} [although he works with a
nonlinear extension to account for ties]; \cite{allison1994} present
related work for ranking statistics in \proglang{SAS} and
\cite{maystre2022} has released a \proglang{python} package for
Luce-type choice datasets.

However, the majority of software is written in the \proglang{R}
computer language~\citep{rcore2023}, which includes extensive
functionality for working with such likelihood functions:
\cite{turner2020} discuss several implementations from a computational
perspective.  The \pkg{BradleyTerry} package \citep{firth2005}
considers pairwise comparisons using \proglang{glm} but cannot deal
with ties or player-specific predictors; the \pkg{BradleyTerry2}
package \citep{turner2012} implements a flexible user interface and
wider range of models to be fitted to pairwise comparison datasets,
specifically simple random effects.  The \pkg{PlackettLuce} package
\citep{turner2020} generalizes this to likelihood functions of the
form of Equation~\ref{PL_like} and applies the Poisson transformation
of~\cite{baker1994} to express the problem as a log-linear model.  The
\pkg{hyper2} package, in contrast, gives a consistent language
interface to create and manipulate likelihood functions over the
simplex ${S}_n=\left\lbrace\left(p_1,\ldots,p_n\right)\left|p_i\geq
0,\sum p_i=1\right.\right\rbrace$.  A further extension in the package
generalizes this likelihood function to functions of ${\mathbf
p}=(p_1,\ldots,p_n)$ with

\begin{equation}\label{hyper2likelihood}
\mathcal{L}\left(\mathbf{p}\right)=
\prod_{s\in \mathcal{O}}\left({\sum_{i\in s}}p_i\right)^{n_s}
\end{equation}

\noindent where~$\mathcal{O}$ is a set of observations and~$s$ a
subset of~$\left\{1,2,\ldots,n\right\}$; numbers~$n_s$ are integers
which may be positive or negative.  The \pkg{hyper2} package has the
ability to evaluate such likelihood functions at any point in $S_n$,
thereby admitting a wide range of non-standard nulls such as order
statistics on the $p_i$~\citep{hankin2017}.  It becomes possible to
analyse a wider range of likelihood functions than standard
Plackett-Luce~\citep{turner2020}.  For example, results involving
incomplete order statistics or teams are tractable.  Further, the
introduction of reified entities (monsters) allows one to consider
draws~\citep{hankin2009}, noncompetitive tactics such as
collusion~\citep{hankin2020}, and the phenomenon of team cohesion
wherein the team becomes stronger than the sum of its
parts~\citep{hankin2010}.  Recent versions of the package include
experimental functionality (\code{cheering()}) to investigate the
relaxing of the assumption of conditional independence of the
forward-ranking process.

Here I present a different generalization.  Consider a race in which
there are six runners 1-6 but we happen to know that three of the
runners (1,2,3) are clones of strength $p_a$, two of the runners (4,5)
have strength $p_b$, and the final runner (6) is of strength $p_c$.
We normalise so $p_a+p_b+p_c=1$.  The runners race and the finishing
order is:

$$a\succ c\succ b\succ a\succ a \succ b$$

Thus the winner was $a$, second place for $c$, third for $b$, and so
on.  Alternatively we might say that $a$ came first, fourth, and
fifth; $b$ came third and sixth, and $c$ came second.  The
Plackett-Luce likelihood function for $p_a,p_b,p_c$ would be

\begin{equation}\label{plackettluce}
\frac{p_a}{3p_a+2p_b+p_c}\cdot
\frac{p_c}{2p_a+2p_b+p_c}\cdot
\frac{p_b}{2p_a+2p_b    }\cdot
\frac{p_a}{2p_a+ p_b    }\cdot
\frac{p_a}{ p_a+ p_b    }\cdot
\frac{p_b}{      p_b    },\qquad p_a+p_b+p_c=1.
\end{equation}

Here I consider such generalized Plackett-Luce likelihood functions,
and give an exact analysis of several simple cases.  I then show how
this class of likelihood functions may be applied to a range of
inference problems involving order statistics.  Illustrative examples,
drawn from Formula 1 motor racing, and track-and-field athletics, are
given.

@@

Existing \code{hyper2} formalism as described by \citet{hankin2017}
cannot represent Equation~\ref{plackettluce}, because
Equation~\ref{hyper2likelihood} uses sets as the indexing elements,
and in this case we need multisets\footnote{Note that the version of
\code{hyper2} presented by \citet{hankin2017} and reviewed
by~\cite{turner2020} used integer-valued sets together with a print
method that used a complicated mapping system from integers to
competitor names.  Current methodology [following commit
  \code{51a8b46}] is to use sets of character strings which represent
the competitors directly; this allows for easier combination of
observations including different competitors.}.  The declarations

\begin{verbatim}
typedef map<string, long double> weightedplayervector;
typedef map<weightedplayervector, long double> hyper3;
\end{verbatim}

show how the \proglang{map} class of the Standard Template Library is
used with \proglang{weightedplayervector} objects mapping strings to
long doubles (specifically, mapping player names to their
multiplicities), and objects of class \proglang{hyper3} are maps from
a \proglang{weightedplayervector} object to long doubles.  One
advantage of this is efficiency: Search, removal, and insertion
operations have logarithmic complexity.  As an example, the following
\proglang{C++} pseudo code would create a log-likelihood function for
the first term in Equation~\ref{plackettluce}:

\begin{verbatim}
weightedplayervector n,d;
n["a"] = 1; 
d["a"] = 3; 
d["b"] = 2;
d["c"] = 1;

hyper3 L;
L[n] = 1;
L[d] = -1;
\end{verbatim}

Above, we understand \code{n} and \code{d} to represent numerator and
denominator respectively.  Object \code{L} is an object of class
\code{hyper3}; it may be evaluated at points in probability space
[that is, a vector \code{[a,b,c]} of nonnegative values with unit sum]
using standard \proglang{R} idiom wrapping \proglang{C++} back end.

@@

The package includes an \proglang{S3} class \code{hyper3} for this
type of object; extraction and replacement methods use \pkg{disordR}
discipline~\citep{hankin2022}.  Package idiom for creating such
objects uses named vectors:

\begin{Schunk}
\begin{Sinput}
R> LL <- hyper3()
R> LL[c(a = 1)] <- 1
R> LL[c(a = 3, b = 2, c = 1)] <- -1
R> LL
\end{Sinput}
\begin{Soutput}
log( (a=1)^1 * (a=3, b=2, c=1)^-1)
\end{Soutput}
\end{Schunk}

Above, we see object \code{LL} is a log-likelihood function of the
players' strengths, which may be evaluated at specified points in
probability space.  A typical use-case would be to assess $H_1\colon
p_a=0.9,p_b=0.05,p_c=0.05$ and $H_2\colon p_a=0.01,p_b=0.01,p_c=0.98$,
and we may evaluate these hypotheses using generic function
\code{loglik()}:

\begin{Schunk}
\begin{Sinput}
R> loglik(c(a = 0.01, b = 0.01, c = 0.98), LL)
\end{Sinput}
\begin{Soutput}
[1] -4.634729
\end{Soutput}
\begin{Sinput}
R> loglik(c(a = 0.90, b = 0.05, c = 0.05), LL)
\end{Sinput}
\begin{Soutput}
[1] -1.15268
\end{Soutput}
\end{Schunk}

Thus we prefer $H_1$ over $H_2$ with about 3.5 units of support,
satisfying the standard two units of support
criterion~\citep{edwards1972}, and we conclude that our observation
[in this case, that one of the three clones of player $a$ beat the $b$
twins and the singleton $c$] furnishes strong support against $H_2$ in
favour of $H_1$.

The package includes many helper functions to work with order
statistics of this type.  Function \code{ordervec2supp3()}, for
example, can be used to generate a log-likelihood function for
Equation~\ref{PL_like}:

\begin{Schunk}
\begin{Sinput}
R> (H <- ordervec2supp3(c("a", "c", "b", "a", "a", "b")))
\end{Sinput}
\begin{Soutput}
log( (a=1)^3 * (a=1, b=1)^-1 * (a=2, b=1)^-1 * (a=2, b=2)^-1 * (a=2,
b=2, c=1)^-1 * (a=3, b=2, c=1)^-1 * (b=1)^1 * (c=1)^1)
\end{Soutput}
\end{Schunk}

(the package gives extensive documentation at
\code{ordervec2supp.Rd}).  We may find a maximum likelihood estimate
for the players' strengths, using generic function \code{maxp()},
dispatching to a specialist \code{hyper3} method:

\begin{Schunk}
\begin{Sinput}
R> (mH <- maxp(H))
\end{Sinput}
\begin{Soutput}
         a          b          c 
0.21324090 0.08724824 0.69951086 
\end{Soutput}
\end{Schunk}

(function \code{maxp()} uses standard optimization techniques to
locate the evaluate; it has access to first derivatives of the
log-likelihood and as such has rapid convergence, if its objective
function is reasonably smooth).

The package provides a number of statistical tests on likelihood
functions, modified from~\citet{hankin2017} to work with \code{hyper3}
objects.  For example, we may assess the hypothesis that all three
players are of equal strength [viz $H_0\colon
p_a=p_b=p_c=\frac{1}{3}$]:

\begin{Schunk}
\begin{Sinput}
R> equalp.test(H)
\end{Sinput}
\begin{Soutput}
	Constrained support maximization

@@

degrees of freedom: 2
support difference = 0.8471613
p-value: 0.42863 
\end{Soutput}
\end{Schunk}

showing, perhaps unsurprisingly, that this small dataset is consistent
with $H_0$.

@@

Arithmetic operations are implemented for \code{hyper3} objects in
much the same way as for \code{hyper2} objects: independent
observations may be combined using the overloaded \code{+} operator;
an example is given below.

The original motivation for \code{hyper3} was the analysis of Formula
1 motor racing, and the package accordingly includes wrappers for
\code{ordervec2supp()} such as \code{ordertable2supp3()} and
\code{attemptstable2supp3()} which facilitate the analysis of commonly
encountered result formats.  Package documentation for order tables is
given at \code{ordertable.Rd} and an example is given below.

@@

Here I consider some order statistics with nontrivial maximum
likelihood Bradley-Terry strengths.  The simplest nontrivial case
would be three competitors with strengths $a,a,b$ and finishing order
$a\succ b\succ a$.  The Plackett-Luce likelihood function would be

\begin{equation}\label{aba}
\frac{a}{2a+b}\cdot\frac{b}{a+b}
\end{equation}

and in this case we know that $a+b=1$ so this is equal to
$\mathcal{L}=\mathcal{L}(a)=\frac{a(1-a)}{1+a}$.  The score would be
given by

\begin{equation}\label{aba_mle}
\frac{d\mathcal{L}}{da}=\frac{(1+a)(1-2a)-a(1-a)}{(1+a)^2}=
\frac{1-2a-a^2}{(1+a)^2}
\end{equation}

and this will be zero at $\sqrt{2}-1$; we also note that
$d^2\mathcal{L}/da^2=-4(1+a)^{-3}$, manifestly strictly negative for
$0\leq a\leq 1$: the root is a maximum.

\begin{Schunk}
\begin{Sinput}
R> maxp(ordervec2supp3(c("a", "b", "a")))
\end{Sinput}
\begin{Soutput}
        a         b 
0.4142108 0.5857892 
\end{Soutput}
\end{Schunk}

Above, we see close agreement with the theoretical value of
$(\sqrt{2}-1,2-\sqrt{2})\simeq (0.414,0.586)$.  Observe that the
maximum likelihood estimate for $a$ is strictly less than 0.5, even
though the finishing order is symmetric.  Using
$\mathcal{L}(a)=\frac{a(1-a)}{1+a}$, we can show that
$\log\mathcal{L}(\hat{a})=\log\left(3-2\sqrt{2}\right)\simeq -1.76$,
where $\hat{a}=\sqrt{2}-1$ is the maximum likelihood estimate for $a$.
Defining $\mathcal{S}=\log\mathcal{L}$ as the support [log-likelihood]
we have

\begin{equation}
\mathcal{S}=\mathcal{S}(a)=\log\left(\frac{a(1-a)}{1+a}\right)-\log\left(3-2\sqrt{2}\right)
\end{equation}

as a standard support function which has a maximum value of zero when
evaluated at $\hat{a}=\sqrt{2}-1$.  For example, we can test the null
that $a=b=\frac{1}{2}$, the statement that the competitors have equal
Bradley-Terry strengths:

\begin{Schunk}
\begin{Sinput}
R> a <- 1/2 # null
R> (S_delta <- log(a * (1 - a)/(1 + a)) - log(3 - 2 * sqrt(2)))
\end{Sinput}
\begin{Soutput}
[1] -0.0290123
\end{Soutput}
\end{Schunk}

Thus the additional support gained in moving from $a=\frac{1}{2}$ to
the evaluate of $a=\sqrt{2}-1$ is 0.029, rather small [as might be
expected given that we have only one rather uninformative observation,
and also given that the maximum likelihood estimate ($\simeq 0.41$) is
quite close to the null of $0.5$].  Nevertheless we can follow
\cite{edwards1972} and apply Wilks's theorem for a $p$~value:

\begin{Schunk}
\begin{Sinput}
R> pchisq(-2 * S_delta, df = 1, lower.tail = FALSE)
\end{Sinput}
\begin{Soutput}
[1] 0.8096458
\end{Soutput}
\end{Schunk}

The $p$~value is about 0.81, exceeding 0.05; thus we have no strong
evidence to reject the null of $a=\frac{1}{2}$.  The observation is
informative, in the sense that we can find a credible interval for
$a$.  With an $n$-units of support criterion the analytical solution to
$\mathcal{S}(p)=-n$ is given by defining $X=\log(3-2\sqrt{2})-n$ and
solving $p(1-p)/(1+p)=X$, or
$p_\pm=\left(1-X\pm\sqrt{1+4X+X^2}\right)/2$, the two roots being the
lower and upper limits of the credible interval;
Figure~\ref{ABA_likelihood}.

\begin{figure}[htbp]
  \begin{center}
\begin{Schunk}
\begin{Sinput}
R> a <- seq(from = 0, by = 0.005, to = 1)
R> S <- function(a){log(a * (1 - a) / ((1 + a) * (3 - 2 * sqrt(2))))}
R> plot(a, S(a), type = 'b',xlab=expression(p[a]),ylab="support")
R> abline(h = c(0, -2))
R> abline(v = c(0.02438102, 0.9524271), col = 'red')
R> abline(v = sqrt(2) - 1)
\end{Sinput}
\end{Schunk}
\includegraphics{jss_hyper3-figaba}
\caption{A\label{ABA_likelihood} support function for $p_a$ with observation $a\succ b\succ a$.}
\end{center}
\end{figure}

\subsubsection{Fisher information}
If we have two clones of $a$ and a singleton $b$, then there are three
possible rank statistics: (i), $a\succ a\succ b$ with probability
$\frac{2a^2}{1+a}$; (ii), $a\succ b\succ a$, with
$\frac{2a(1-a)}{(1+a)}$, (iii), $b\succ a\succ a$ at
$\frac{1-a}{1+a}$.  Likelihood functions for these order statistics
are given in Figure~\ref{ABAetc_likelihood}.  It can be shown that the
Fisher information for such observations is

\begin{equation}
\mathcal{I}(a)=2\frac{1+a+a^2}{a(1-a)(1+a)^2}
\end{equation}

which has a minimum of about $6.21$ at at about $a=0.522$.  We can
compare this with the Fisher information matrix ${\mathcal I}$, for
the case of three distinct runners $a,b,c$, evaluated at its minimum
of $p_a=p_b=p_c=\frac{1}{3}$.  If we observe the complete order
statistic, $\left|{\mathcal I}\right| =\frac{1323}{16}\simeq 82.7$; if
we observe just the winner, $\left|{\mathcal I}\right|=27$, and if we
observe just the loser we have $\left|{\mathcal
I}\right|=\frac{16875}{256}\simeq 65.9$.  A brief discussion is given
at \code{inst/fisher\_inf\_PL3.Rmd}.

>>>>

\usepackage{orcidlink,thumbpdf,lmodern}
\graphicspath{{Figures/}}

@@

%% an abstract and keywords
\Abstract{
  The \pkg{hyper2} package provides functionality to work with
  extensions of the Bradley-Terry probability model such as
  Plackett-Luce likelihood including team strengths and reified
  entities (monsters).  The package allows one to use relatively
  natural \proglang{R}~idiom to manipulate such likelihood functions.  Here, I
  present a generalization of \code{hyper2} in which multiple entities
  are constrained to have identical Bradley-Terry strengths.  A new
  \proglang{S3} class `\code{hyper3}', along with associated methods, is
  motivated and introduced.  Three datasets are analysed, each
  analysis furnishing new insight, and each highlighting different
  capabilities of the package.  }

\Keywords{Plackett-Luce, Bradley-Terry, Mann-Whitney}
\Plainkeywords{Plackett-Luce, Bradley-Terry, Mann-Whitney}

@@

The \pkg{hyper2} package \citep{hankin2009,hankin2017} furnishes
computational support for generalized
Plackett-Luce \citep{plackett1975} likelihood functions.  The
preferred interpretation is a race (as in track and field athletics):
Given six competitors $1-6$, we ascribe to them nonnegative strengths
$p_1,\ldots, p_6$; the probability that $i$ beats $j$ is
$p_i/(p_i+p_j)$.  It is conventional to normalize so that the total
strength is unity, and to identify a competitor with his strength.
Given an order statistic, say $p_1\succ p_2\succ p_3\succ p_4$, the
Plackett-Luce likelihood function would be
%
\begin{equation}\label{PL_like}
  \frac{p_1}{p_1+p_2+p_3+p_4}\cdot
  \frac{p_2}{    p_2+p_3+p_4}\cdot
  \frac{p_3}{        p_3+p_4}\cdot
  \frac{p_4}{            p_4}.
\end{equation}
%
\citet{mollica2014} call this a forward ranking process on the grounds
that the best (preferred; fastest; chosen) entities are identified in
the same sequence as their rank.

Computational support for Bradley-Terry likelihood functions is
available in a range of languages.  \cite{hunter2004}, for example,
presents results in \proglang{MATLAB} (although he works with a
nonlinear extension to account for ties); \cite{allison1994} present
related work for ranking statistics in \proglang{SAS} and
\cite{maystre2022} has released a \proglang{Python} package for
Luce-type choice datasets.

However, the majority of software is written in the \proglang{R}~computer language \citep{rcore2023}, which includes extensive
functionality for working with such likelihood functions:
\cite{turner2020} discuss several implementations from a computational
perspective.  The \pkg{BradleyTerry} package \citep{firth2005}
considers pairwise comparisons using \code{glm} but cannot deal
with ties or player-specific predictors; the \pkg{BradleyTerry2}
package \citep{turner2012} implements a flexible user interface and
wider range of models to be fitted to pairwise comparison datasets,
specifically simple random effects.  The \pkg{PlackettLuce} package
\citep{turner2020} generalizes this to likelihood functions of the
form of Equation~\ref{PL_like} and applies the Poisson transformation
of~\cite{baker1994} to express the problem as a log-linear model.  The
\pkg{hyper2} package, in contrast, gives a consistent language
interface to create and manipulate likelihood functions over the
simplex ${S}_n=\left\lbrace\left(p_1,\ldots,p_n\right)\mid p_i\geq
0,\sum p_i=1\right\rbrace$.  A further extension in the package
generalizes this likelihood function to functions of ${\mathbf
p}=(p_1,\ldots,p_n)$ with
%
\begin{equation}\label{hyper2likelihood}
\mathcal{L}\left(\mathbf{p}\right)=
\prod_{s\in \mathcal{O}}\left({\sum_{i\in s}}p_i\right)^{n_s}
\end{equation}
%
where~$\mathcal{O}$ is a set of observations and~$s$ a
subset of~$\left\{1,2,\ldots,n\right\}$; numbers~$n_s$ are integers
which may be positive or negative.  The \pkg{hyper2} package has the
ability to evaluate such likelihood functions at any point in $S_n$,
thereby admitting a wide range of non-standard nulls such as order
statistics on the $p_i$ \citep{hankin2017}.  It becomes possible to
analyse a wider range of likelihood functions than standard
Plackett-Luce \citep{turner2020}.  For example, results involving
incomplete order statistics or teams are tractable.  Further, the
introduction of reified entities (monsters) allows one to consider
draws \citep{hankin2009}, noncompetitive tactics such as
collusion \citep{hankin2020}, and the phenomenon of team cohesion
wherein the team becomes stronger than the sum of its
parts \citep{hankin2010}.  Recent versions of the package include
experimental functionality (\code{cheering()}) to investigate the
relaxing of the assumption of conditional independence of the
forward-ranking process.

Here I present a different generalization.  Consider a race in which
there are six runners 1-6 but we happen to know that three of the
runners (1, 2, 3) are clones of strength $p_a$, two of the runners (4, 5)
have strength $p_b$, and the final runner (6) is of strength $p_c$.
We normalize so $p_a+p_b+p_c=1$.  The runners race and the finishing
order is:
%
$$a\succ c\succ b\succ a\succ a \succ b$$
%
Thus the winner was $a$, second place for $c$, third for $b$, and so
on.  Alternatively we might say that $a$ came first, fourth, and
fifth; $b$ came third and sixth, and $c$ came second.  The
Plackett-Luce likelihood function for $p_a, p_b, p_c$ would be
%
\begin{equation}\label{plackettluce}
\frac{p_a}{3p_a+2p_b+p_c}\cdot
\frac{p_c}{2p_a+2p_b+p_c}\cdot
\frac{p_b}{2p_a+2p_b    }\cdot
\frac{p_a}{2p_a+ p_b    }\cdot
\frac{p_a}{ p_a+ p_b    }\cdot
\frac{p_b}{      p_b    },\qquad p_a+p_b+p_c=1.
\end{equation}
%
Here I consider such generalized Plackett-Luce likelihood functions,
and give an exact analysis of several simple cases.  I then show how
this class of likelihood functions may be applied to a range of
inference problems involving order statistics.  Illustrative examples,
drawn from Formula 1 motor racing, and track-and-field athletics, are
given.

@@

Existing \code{hyper2} formalism as described by \citet{hankin2017}
cannot represent Equation~\ref{plackettluce}, because
Equation~\ref{hyper2likelihood} uses sets as the indexing elements,
and in this case we need multisets\footnote{Note that the version of
\code{hyper2} presented by \citet{hankin2017} and reviewed
by~\cite{turner2020} used integer-valued sets together with a print
method that used a complicated mapping system from integers to
competitor names.  Current methodology (following commit
  \code{51a8b46}) is to use sets of character strings which represent
the competitors directly; this allows for easier combination of
observations including different competitors.}.  The declarations
%
\begin{verbatim}
typedef map<string, long double> weightedplayervector;
typedef map<weightedplayervector, long double> hyper3;
\end{verbatim}
%
show how the `\code{map}' class of the Standard Template Library is
used with `\code{weightedplayervector}' objects mapping strings to
long doubles (specifically, mapping player names to their
multiplicities), and objects of class `\code{hyper3}' are maps from
a `\code{weightedplayervector}' object to long doubles.  One
advantage of this is efficiency: Search, removal, and insertion
operations have logarithmic complexity.  As an example, the following
\proglang{C++} pseudo code would create a log-likelihood function for
the first term in Equation~\ref{plackettluce}:
%
\begin{verbatim}
weightedplayervector n,d;
n["a"] = 1; 
d["a"] = 3; 
d["b"] = 2;
d["c"] = 1;

hyper3 L;
L[n] = 1;
L[d] = -1;
\end{verbatim}
%
Above, we understand \code{n} and \code{d} to represent numerator and
denominator respectively.  Object \code{L} is an object of class
`\code{hyper3}'; it may be evaluated at points in probability space
(that is, a vector \code{[a,b,c]} of nonnegative values with unit sum)
using standard \proglang{R} idiom wrapping \proglang{C++} back end.

@@

The package includes an \proglang{S}3 class `\code{hyper3}' for this
type of object; extraction and replacement methods use \pkg{disordR}
discipline \citep{hankin2022}.  Package idiom for creating such
objects uses named vectors:
%
\begin{Schunk}
\begin{Sinput}
R> LL <- hyper3()
R> LL[c(a = 1)] <- 1
R> LL[c(a = 3, b = 2, c = 1)] <- -1
R> LL
\end{Sinput}
\begin{Soutput}
log( (a=1)^1 * (a=3, b=2, c=1)^-1)
\end{Soutput}
\end{Schunk}
%
Above, we see object \code{LL} is a log-likelihood function of the
players' strengths, which may be evaluated at specified points in
probability space.  A typical use-case would be to assess $H_1\colon
p_a=0.9,p_b=0.05,p_c=0.05$ and $H_2\colon p_a=0.01,p_b=0.01,p_c=0.98$,
and we may evaluate these hypotheses using generic function
\code{loglik()}:
%
\begin{Schunk}
\begin{Sinput}
R> loglik(c(a = 0.01, b = 0.01, c = 0.98), LL)
\end{Sinput}
\begin{Soutput}
[1] -4.634729
\end{Soutput}
\begin{Sinput}
R> loglik(c(a = 0.90, b = 0.05, c = 0.05), LL)
\end{Sinput}
\begin{Soutput}
[1] -1.15268
\end{Soutput}
\end{Schunk}
%
Thus we prefer $H_1$ over $H_2$ with about 3.5 units of support,
satisfying the standard two units of support
criterion \citep{edwards1972}, and we conclude that our observation
[in this case, that one of the three clones of player $a$ beat the $b$
twins and the singleton $c$] furnishes strong support against $H_2$ in
favor of $H_1$.

The package includes many helper functions to work with order
statistics of this type.  Function \code{ordervec2supp3()}, for
example, can be used to generate a log-likelihood function for
Equation~\ref{PL_like}:
%
\begin{Schunk}
\begin{Sinput}
R> (H <- ordervec2supp3(c("a", "c", "b", "a", "a", "b")))
\end{Sinput}
\begin{Soutput}
log( (a=1)^3 * (a=1, b=1)^-1 * (a=2, b=1)^-1 * (a=2, b=2)^-1 * (a=2,
b=2, c=1)^-1 * (a=3, b=2, c=1)^-1 * (b=1)^1 * (c=1)^1)
\end{Soutput}
\end{Schunk}
%
(the package gives extensive documentation at
\code{ordervec2supp.Rd}).  We may find a maximum likelihood estimate
for the players' strengths, using generic function \code{maxp()},
dispatching to a specialist \code{hyper3} method:
%
\begin{Schunk}
\begin{Sinput}
R> (mH <- maxp(H))
\end{Sinput}
\begin{Soutput}
         a          b          c 
0.21324090 0.08724824 0.69951086 
\end{Soutput}
\end{Schunk}
%
(function \code{maxp()} uses standard optimization techniques to
locate the evaluate; it has access to first derivatives of the
log-likelihood and as such has rapid convergence, if its objective
function is reasonably smooth).

The package provides a number of statistical tests on likelihood
functions, modified from~\citet{hankin2017} to work with `\code{hyper3}'
objects.  For example, we may assess the hypothesis that all three
players are of equal strength (viz $H_0\colon
p_a=p_b=p_c=\frac{1}{3}$):
%
\begin{Schunk}
\begin{Sinput}
R> equalp.test(H)
\end{Sinput}
\begin{Soutput}
	Constrained support maximization

@@

degrees of freedom: 2
support difference = 0.8471613
p-value: 0.42863 
\end{Soutput}
\end{Schunk}
%
showing, perhaps unsurprisingly, that this small dataset is consistent
with $H_0$.

@@

Arithmetic operations are implemented for `\code{hyper3}' objects in
much the same way as for `\code{hyper2}' objects: independent
observations may be combined using the overloaded \code{+} operator;
an example is given below.

The original motivation for \code{hyper3} was the analysis of
Formula~1 motor racing, and the package accordingly includes wrappers for
\code{ordervec2supp()} such as \code{ordertable2supp3()} and
\code{attemptstable2supp3()} which facilitate the analysis of commonly
encountered result formats.  Package documentation for order tables is
given at \code{ordertable.Rd} and an example is given below.

@@

Here I consider some order statistics with nontrivial maximum
likelihood Bradley-Terry strengths.  The simplest nontrivial case
would be three competitors with strengths $a,a,b$ and finishing order
$a\succ b\succ a$.  The Plackett-Luce likelihood function would be
%
\begin{equation}\label{aba}
\frac{a}{2a+b}\cdot\frac{b}{a+b}
\end{equation}
%
and in this case we know that $a+b=1$ so this is equal to
$\mathcal{L}=\mathcal{L}(a)=\frac{a(1-a)}{1+a}$.  The score would be
given by
%
\begin{equation}\label{aba_mle}
\frac{d\mathcal{L}}{da}=\frac{(1+a)(1-2a)-a(1-a)}{(1+a)^2}=
\frac{1-2a-a^2}{(1+a)^2}
\end{equation}
%
and this will be zero at $\sqrt{2}-1$; we also note that
$d^2\mathcal{L}/da^2=-4(1+a)^{-3}$, manifestly strictly negative for
$0\leq a\leq 1$: the root is a maximum.
%
\begin{Schunk}
\begin{Sinput}
R> maxp(ordervec2supp3(c("a", "b", "a")))
\end{Sinput}
\begin{Soutput}
        a         b 
0.4142108 0.5857892 
\end{Soutput}
\end{Schunk}
%
Above, we see close agreement with the theoretical value of
$(\sqrt{2}-1,2-\sqrt{2})\simeq (0.414,0.586)$.  Observe that the
maximum likelihood estimate for $a$ is strictly less than 0.5, even
though the finishing order is symmetric.  Using
$\mathcal{L}(a)=\frac{a(1-a)}{1+a}$, we can show that
$\log\mathcal{L}(\hat{a})=\log\left(3-2\sqrt{2}\right)\simeq -1.76$,
where $\hat{a}=\sqrt{2}-1$ is the maximum likelihood estimate for $a$.
Defining $\mathcal{S}=\log\mathcal{L}$ as the support [log-likelihood]
we have
%
\begin{equation}
\mathcal{S}=\mathcal{S}(a)=\log\left(\frac{a(1-a)}{1+a}\right)-\log\left(3-2\sqrt{2}\right)
\end{equation}
%
as a standard support function which has a maximum value of zero when
evaluated at $\hat{a}=\sqrt{2}-1$.  For example, we can test the null
that $a=b=\frac{1}{2}$, the statement that the competitors have equal
Bradley-Terry strengths:
%
\begin{Schunk}
\begin{Sinput}
R> a <- 1/2
R> (S_delta <- log(a * (1 - a)/(1 + a)) - log(3 - 2 * sqrt(2)))
\end{Sinput}
\begin{Soutput}
[1] -0.0290123
\end{Soutput}
\end{Schunk}
%
Thus the additional support gained in moving from $a=\frac{1}{2}$ to
the evaluate of $a=\sqrt{2}-1$ is 0.029, rather small [as might be
expected given that we have only one rather uninformative observation,
and also given that the maximum likelihood estimate ($\simeq 0.41$) is
quite close to the null of $0.5$].  Nevertheless we can follow
\cite{edwards1972} and apply Wilks's theorem for a $p$~value:
%
\begin{Schunk}
\begin{Sinput}
R> pchisq(-2 * S_delta, df = 1, lower.tail = FALSE)
\end{Sinput}
\begin{Soutput}
[1] 0.8096458
\end{Soutput}
\end{Schunk}
%
The $p$~value is about 0.81, exceeding 0.05; thus we have no strong
evidence to reject the null of $a=\frac{1}{2}$.  The observation is
informative, in the sense that we can find a credible interval for
$a$.  With an $n$-units of support criterion the analytical solution to
$\mathcal{S}(p)=-n$ is given by defining $X=\log(3-2\sqrt{2})-n$ and
solving $p(1-p)/(1+p)=X$, or
$p_\pm=\left(1-X\pm\sqrt{1+4X+X^2}\right)/2$, the two roots being the
lower and upper limits of the credible interval;
Figure~\ref{ABA_likelihood}.
%
\begin{Schunk}
\begin{Sinput}
R> a <- seq(from = 0, by = 0.005, to = 1)
R> S <- function(a){log(a * (1 - a) / ((1 + a) * (3 - 2 * sqrt(2))))}
R> plot(a, S(a), type = "b", xlab = expression(p[a]), ylab = "support")
R> abline(h = c(0, -2))
R> abline(v = c(0.02438102, 0.9524271), col = "red")
R> abline(v = sqrt(2) - 1)
\end{Sinput}
\end{Schunk}
%
\begin{figure}[t!]
\centering
\includegraphics[trim = 0 5 0 20,clip]{jss_hyper3-figaba}
\caption{A\label{ABA_likelihood} support function for $p_a$ with observation $a\succ b\succ a$.}
\end{figure}

\subsubsection{Fisher information}

If we have two clones of $a$ and a singleton $b$, then there are three
possible rank statistics: (i), $a\succ a\succ b$ with probability
$\frac{2a^2}{1+a}$; (ii), $a\succ b\succ a$, with
$\frac{2a(1-a)}{(1+a)}$, (iii), $b\succ a\succ a$ at
$\frac{1-a}{1+a}$.  Likelihood functions for these order statistics
are given in Figure~\ref{ABAetc_likelihood}.  It can be shown that the
Fisher information for such observations is
%
\begin{equation}
\mathcal{I}(a)=2\frac{1+a+a^2}{a(1-a)(1+a)^2}
\end{equation}
%
which has a minimum of about $6.21$ at at about $a=0.522$.  We can
compare this with the Fisher information matrix ${\mathcal I}$, for
the case of three distinct runners $a,b,c$, evaluated at its minimum
of $p_a=p_b=p_c=\frac{1}{3}$.  If we observe the complete order
statistic, $\left|{\mathcal I}\right| =\frac{1323}{16}\simeq 82.7$; if
we observe just the winner, $\left|{\mathcal I}\right|=27$, and if we
observe just the loser we have $\left|{\mathcal
I}\right|=\frac{16875}{256}\simeq 65.9$.  A brief discussion is given
at \code{inst/fisher\_inf\_PL3.Rmd}.

\begin{figure}[t!]
\centering
\includegraphics[trim = 0 5 0 20,clip]{jss_hyper3-figabbabb}
\caption{Likelihood \label{ABAetc_likelihood} functions for
observations $a\succ a\succ b$, $a\succ b\succ a$, $b\succ a\succ a$.  Horizontal dotted
line represents two units of support}
\end{figure}
